---
title: "Bootstrap weights"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bootstrap weights}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = nzchar(Sys.getenv("COMPILE_VIG_CANPUMF"))
)
```

```{r setup, warning=FALSE}
library(dplyr)
library(canpumf)
options(canpumf.cache_path = Sys.getenv("COMPILE_VIG_CANPUMF"))
```

PUMF data are a synthetic sample, altered from the original survey responses for
privacy reasons. As with any survey sample, point estimates carry sampling
uncertainty, and it is good practice to quantify it. Some PUMF ship with
bootstrap weights for exactly this purpose; when they do not, **canpumf** can
generate them with `add_bootstrap_weights()`.

The weights a survey ships with need no extra step: they are columns of the
table `get_pumf()` returns (`BSW1`, ..., `BSW1000` for the Survey of Financial
Security, `WT1`, ..., `WT16` for the Census). `bsw_info()` lists them as
`source = "survey"`, next to the `source = "generated"` weights this vignette is
about:

```{r, eval = FALSE}
get_pumf("SFS", "2019") |> bsw_info()
#> # A tibble: 1 x 7
#>   source weight_col prefix bsw_table     temporary n_replicates size_mb
#>   <chr>  <chr>      <chr>  <chr>         <lgl>            <int>   <dbl>
#> 1 survey NA         BSW_   eng_EFAM_PUMF FALSE             1000    83.4
```

This vignette explains how `add_bootstrap_weights()` works: the method it uses,
where the weights live, how stratification works, and — importantly — what
happens when you call it again on a dataset that already has weights.

## The method: a resampling bootstrap

For each of `n_replicates` replicates, a sample of `n` rows is drawn **with
replacement** from the `n` rows of the (stratum of the) survey. The bootstrap
weight for row $i$ in replicate $b$ is

$$ w^{*}_{i,b} = w_i \cdot c_{i,b}, $$

where $w_i$ is the original survey weight and $c_{i,b}$ is the number of times
row $i$ was selected in replicate $b$. Because each row is selected once on
average, $\mathbb{E}[c_{i,b}] = 1$ and the replicate weights are unbiased for
the original weights. The spread of an estimate across the replicates is what
gives you its sampling variance (see [Estimating uncertainty](#estimating-uncertainty)).

The default is `n_replicates = 500`. Pass `seed` for reproducibility.

## Two modes

`add_bootstrap_weights()` works on two kinds of input and behaves differently
for each:

* **A DuckDB-backed lazy table** returned by `get_pumf()` (the typical case).
  The replicate weights go into a table of their own, keyed like the survey
  table, and the function returns the input table joined to it. Whether that
  table is stored in the DuckDB file or is temporary depends on how the survey
  table was opened (next section). Stored weights are **reused by subsequent
  calls** instead of being recomputed.
* **An in-memory `data.frame` / `tibble`** (e.g. after `collect()`). The
  weights are generated in memory and the augmented data frame is returned.
  Nothing is persisted.

```{r}
cis <- get_pumf("CIS", "2019")

cis_bsw <- cis |>
  add_bootstrap_weights(weight_col = "FWEIGHT", n_replicates = 200, seed = 42)

# 200 replicate columns CPBSW1 … CPBSW200 are now available
grep("^CPBSW", colnames(cis_bsw), value = TRUE) |> head()
```

## Where the weights are stored (DuckDB path)

The replicate weights are **not** copied into the main survey table. They are
written to a separate table, auto-named
`paste0("pumf_bsw_", tolower(weight_col))` (e.g. `pumf_bsw_fweight`). The
auto-naming means generating weights for two different weight columns
(e.g. a household weight and a person weight) produces two independent BSW
tables that never collide.

`add_bootstrap_weights()` works on the connection of the table it is given. It
never closes or reopens one, so what it can store depends on that connection:

* **Read-only** (the `get_pumf()` default). Stored weights are used when there
  are any. Otherwise the weights go to a temporary table of the connection
  (`tmp_pumf_bsw_fweight`). The DuckDB file is not touched, other sessions
  reading the same survey are not disturbed, and the weights are gone when the
  connection is closed (`close_pumf()`, or the end of the R session). A message
  says so. Pass `seed` to get the same weights the next time.
* **Read-write** (`get_pumf(..., read_only = FALSE)`). The weights are stored in
  the DuckDB file and reused by every later call, in this session or another,
  and also by read-only tables. A write connection needs the database to
  itself: close the read-only tables of the survey first, and `close_pumf()`
  the write connection when the weights are stored.

```{r, eval = FALSE}
# once: store the weights
close_pumf(cis)     # the read-only table opened above holds the database
cis_rw <- get_pumf("CIS", "2019", read_only = FALSE)
cis_rw |> add_bootstrap_weights(weight_col = "FWEIGHT", n_replicates = 200, seed = 42)
close_pumf(cis_rw)

# from then on, in any session: the stored weights are found and joined
cis_bsw <- get_pumf("CIS", "2019") |>
  add_bootstrap_weights(weight_col = "FWEIGHT", n_replicates = 200)
```

`bsw_info()` summarises the bootstrap weights of the table. `source` tells the
generated weights (`temporary` says whether they are stored) from replicate
weights that came with the survey, which the CIS does not have:

```{r}
bsw_info(cis_bsw)
```

The weights always cover the complete survey table. For a temporary table that
is rows x replicates x 8 bytes, which is small for most surveys and several
gigabytes for the Census or the full LFS; `add_bootstrap_weights()` warns
before it builds a temporary table of more than about 2 GB. Store such weights
through a write connection, or `collect()` the rows of interest and add the
weights to the data frame.

## Identifying rows

To join replicate weights back to the survey rows a key that identifies each
row is needed. With `id_col = NULL` (the default):

* the survey registry's `bsw_join_key` is used when one is defined (e.g.
  `PEFAMID` for the SFS); otherwise
* the `pumf_row_id` column that every table built by canpumf 0.7.0 or later
  carries;
* for the longitudinal series (`"LFS"`, `"LFS_HIST"`), `SURVYEAR`, `SURVMNTH`
  and `REC_NUM` together.

You can always pass `id_col` explicitly to use a natural key, which may consist
of several columns. A table built by an earlier version of canpumf that has
none of these needs a rebuild (`get_pumf(..., refresh = TRUE)`) or an `id_col`.

## Stratified bootstrap weights

Many survey designs are stratified, and the resampling should respect that:
each replicate should resample **within** each stratum, preserving the stratum
sample sizes. Pass `strata_cols`:

```{r, eval = FALSE}
cis_bsw <- cis |>
  add_bootstrap_weights(weight_col = "FWEIGHT", strata_cols = "PROV",
                        n_replicates = 200, seed = 42)
```

Resolution order for the strata:

1. `strata_cols` if you pass it,
2. otherwise the registry `bsw_strata` for the survey,
3. otherwise, for the LFS, the default `c("SURVYEAR", "SURVMNTH")` so each month
   is resampled separately,
4. otherwise unstratified.

Pass `strata_cols = character(0)` to force unstratified weights even when a
default exists (e.g. for the LFS).

## Estimating uncertainty {#estimating-uncertainty}

To get a bootstrap standard error for an estimate, compute the estimate once
with the original weights and once with each replicate weight; the spread of
the replicate estimates is the sampling variability. The example below
estimates the total population represented by the survey (the sum of the
weights), which depends only on the weight columns, but the same pattern
applies to any weighted statistic — replace `sum(.x)` with your estimator
(a weighted mean, share, total, …) evaluated with each weight column.

```{r}
est <- cis_bsw |>
  summarise(across(c(FWEIGHT, matches("^CPBSW[0-9]+$")), ~ sum(.x))) |>
  collect()

point_estimate      <- est$FWEIGHT
replicate_estimates <- est |> select(matches("^CPBSW[0-9]+$")) |> unlist()

# Bootstrap variance: mean squared deviation of the replicate estimates from
# the full-sample estimate; the standard error is its square root.
# Confidence intervals can be obtained by taking the appropriate quantiles.
confidence_interval <- quantile(replicate_estimates,c(0.025,0.975))
std_error <- sqrt(mean((replicate_estimates - point_estimate)^2))

c(estimate = point_estimate, se = std_error, conf=confidence_interval)
```

The same pattern works after `group_by()` / `summarise()`: carry the `CPBSW*`
columns through the summary, then collapse them into a standard error per group.

## Incremental re-runs

Calling `add_bootstrap_weights()` again on a survey that already has weights
does only the work needed to satisfy the request. The decision is driven by
two questions: are there enough replicate columns, and do all rows have
weights?

### Reuse — nothing to do

If the weights already cover every row and there are at least `n_replicates` of
them, they are reused without recomputation (requesting fewer than exist simply
returns a subset). This is why repeatedly opening a survey with stored weights
is cheap. On a read-only connection the temporary weights of that connection
are looked at first, then the stored ones.

### More replicates requested

If you ask for more replicates than exist (and no rows were added), the
additional replicate columns are generated and appended; the existing columns
are kept unchanged. When stratified, the new columns are resampled within
strata, consistent with the existing ones.

```{r, eval = FALSE}
# First 200 replicates …
cis_bsw <- add_bootstrap_weights(cis, "FWEIGHT", n_replicates = 200, seed = 42)
# … later, extend to 500: only CPBSW201 … CPBSW500 are generated.
cis_bsw <- add_bootstrap_weights(cis, "FWEIGHT", n_replicates = 500, seed = 42)
```

### Rows added to the survey table

This is the subtle case. A bootstrap replicate resamples the **full
population** (or full stratum), so adding rows invalidates the replicate
weights of the affected resampling universe — you cannot simply generate
weights for the new rows in isolation. `add_bootstrap_weights()` detects that
not all rows have weights and **deletes and regenerates** the affected weights:

* **Unstratified:** the whole population changed, so every row's replicate
  weights are regenerated.
* **Stratified:** only the strata that gained rows are regenerated in full;
  strata that are already complete keep their existing weights. This makes
  re-weighting an updated multi-year file (for example, appending a new LFS
  month or year) inexpensive — only the new strata are recomputed.

Extending the number of replicates and adding rows can happen in the same call;
each is handled on the rows it applies to.

On a read-only connection both cases are worked out in the temporary table:
the stored replicates are copied into it, so they keep their values, and the
stored weights stay as they are.

> Extending a weighted table after appending rows relies on the new rows having
> a valid identifier. Use a natural `id_col` (or the registry `bsw_join_key`)
> when you expect to grow the table, so the new rows can be matched.

### Forcing a full regeneration

Pass `overwrite = TRUE` to discard the weights and regenerate from scratch
regardless of what is already there: the stored weights on a write connection,
the temporary ones on a read-only connection.

```{r, eval = FALSE}
cis_bsw <- add_bootstrap_weights(cis, "FWEIGHT", n_replicates = 200,
                                 seed = 42, overwrite = TRUE)
```

## Multiple weight columns

Because each BSW table is named after its weight column, hierarchical surveys
with several weights are handled by simply calling the function once per weight.
By default all replicate columns share the `CPBSW` prefix, and a call replaces
the replicate columns of its prefix that the input already carries. Give each
weight column a distinct `prefix` to have both sets of replicates in one table
(the column names are survey-specific):

```{r, eval = FALSE}
tbl <- get_pumf("<series>", "<version>") |>
  add_bootstrap_weights(weight_col = "<household_weight>", prefix = "HHBSW") |>
  add_bootstrap_weights(weight_col = "<person_weight>",    prefix = "PPBSW")
```

## What the input table carries

Bootstrap weights always cover the complete physical survey table, so every
subset is weighted from the same replicates. The result is the input `tbl`
joined to the weights: a `filter()`, `select()`, `mutate()` or
`label_pumf_columns()` applied before the call is still in effect afterwards.

```{r, eval = FALSE}
cis |>
  filter(PROV == "Ontario") |>
  label_pumf_columns() |>
  add_bootstrap_weights(weight_col = "FWEIGHT", n_replicates = 200, seed = 42)
#> the Ontario rows, with labelled columns and CPBSW1 ... CPBSW200
```

Two things to keep in mind:

* The column(s) that identify the rows must still be among the columns of the
  input, since the join runs on them. Add the weights before a `select()` or
  `summarise()` drops them.
* The order of an `arrange()` does not survive a join; sort the result instead.

## Connection lifecycle

`add_bootstrap_weights()`, `bsw_info()` and `remove_bootstrap_weights()` never
close a connection. The table passed in, and every other table on the same
connection, stays valid, and the returned table lives on that connection too.
Temporary weights last exactly as long as it does: after `close_pumf()` they
are gone, and the next call generates them again (or finds stored ones).

## Inspecting and removing weights

```{r, eval = FALSE}
bsw_info(cis_bsw)                      # the survey's own and the generated weights, replicate counts
cis_clean <- remove_bootstrap_weights(cis_bsw, weight_col = "FWEIGHT")
```

`remove_bootstrap_weights()` drops the temporary weights of the connection and
returns a tbl on the original survey table. Stored weights are dropped on a
write connection only; on a read-only connection they are kept. The replicate
weights that came with the survey are part of the survey table and are never
removed.
