---
title: "Census"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Census}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = nzchar(Sys.getenv("COMPILE_VIG_CANPUMF"))
)
```

```{r setup}
library(canpumf)
library(dplyr)
library(ggplot2)
options(canpumf.cache_path = Sys.getenv("COMPILE_VIG_CANPUMF"))
```

The package supports Census PUMF from 1971 through 2021, covering individuals, hierarchical, households, and families variants depending on the year. The 2021 and 2016 files are available via direct download; older years must be ordered through Statistics Canada's EFT portal and placed in the directory pointed to by the `canpumf.cache_path` option.

```{r}
census_2021 <- get_pumf("Census",version="2021") 
```


As a simple applicartion we look at household maintainer rates by age group for select metro areas, using standard weights.

```{r}
census_2021 |>
  filter(CMA %in% c("Vancouver","Toronto","Montréal","Québec")) |>
  filter(PRIHM!="Not applicable") |>
  filter(AGEGRP!="Not available") |>
  summarise(across(matches("WEIGHT|WT\\d+"),sum),
            .by=c(CMA,AGEGRP,PRIHM)) |>
  mutate(Share=WEIGHT/sum(WEIGHT),.by=c(CMA,AGEGRP)) |>
  filter(PRIHM=="Person is primary maintainer") |>
  ggplot(aes(y=AGEGRP,x=Share,fill=CMA)) +
  geom_bar(stat="identity",position="dodge") +
  scale_x_continuous(labels=scales::percent) +
  labs(title="Age-specifc household maintainer rates",
       y="Age group",
       x="Household maintainer rate",
       caption="StatCan Census 2021 PUMF") 
```
