---
title: "Historical Census Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with canpumf}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = nzchar(Sys.getenv("COMPILE_VIG_CANPUMF"))
)
```

```{r setup}
library(dplyr)
library(tidyr)
library(ggplot2)
library(canpumf)
options(canpumf.cache_path = Sys.getenv("COMPILE_VIG_CANPUMF"))
```

There are two sets of historical census data that **canpumf** has registry entry for automatic ingenstion, one is the 1881 full sample curated by the TCP, and the other is the 1991 5% sample curated by CCRI. This vignette gives a quick overview how to work with the data.


These census data come from transcribed scanned handwritten census forms, and they contain problems and might need cleaning and imputation when used for specific purposes.

## TCP 1881 census

```{r}
census_1881 <- get_pumf("TCP","1881")
```

A simple application is to look at the distribution of household sizes. Households are distinguished by DIVNO, DISTNO, SDISTLET, and HHNBR.

```{r}
census_1881 |>
  summarise(hh_size=n(),.by=c(DIVNO, DISTNO, SDISTLET,HHNBR)) |>
  count(hh_size) |>
  ggplot(aes(x=hh_size,y=n, group=hh_size)) +
  geom_col() +
  coord_cartesian(xlim = c(0,20)) +
  scale_x_continuous(breaks=seq(0,150,10)) +
  labs(title="Household size in Canada",
       x="Size of household (scale truncated at 20)",
       y="Number of households",
       caption="Canada census 1881 (TCP full samele")
```


When interested in household level data we might want to filter down to the oldest member in each household so we have one person per household. 

```{r}
census_1881 |>
  slice_max(n=1,order_by=AGE, with_ties=FALSE,
            by=c(DIVNO, DISTNO, SDISTLET,HHNBR)) |>
  count(AGE) |>
  ggplot(aes(x=AGE,y=n, group=AGE)) +
  geom_col() +
  scale_x_continuous(breaks=seq(0,150,10)) +
  labs(title="Age of oldest person in each household",
       x="Age of oldest person in each household",
       y="Number of households",
       caption="Canada census 1881 (TCP full samele")
```

The high number of households with oldest person aged below 20 is quite remarkable and likely a data parsing artefact that is worth investigating further before overly relying on this data for analysis.

The clustering at round ages is interesting and also quite visible when just looking at the overall age distribution.

```{r}
census_1881 |>
  count(AGE,SEX) |>
  mutate(n=if_else(SEX=="Female",-n,n)) |>
  collect() |>
  filter(!is.na(AGE), SEX %in% c("Male","Female")) |>
  ggplot(aes(y=factor(AGE),x=n,fill=as.character(SEX))) +
  geom_col() +
  scale_y_discrete(breaks=seq(0,150,5)) +
  scale_x_continuous(labels=\(x)scales::comma(abs(x))) +
  labs(title="Age pyramid in 1881",
       x="Number of people",y="Age",fill="Sex",
       caption="Canada census 1881 (TCP full samele")
```

```{r}
close_pumf(census_1881)
```


## CCRI 1911 census 5% sample

```{r}
census_1911 <- get_pumf("CCRI","1911")
```

The CCIR data is somewhat messy and requires some imputation.


```{r}
census_1911 |>
  mutate(AGE=case_when(AGE_UNIT=="Day" ~ floor(AGE_AMOUNT/365),
                       AGE_UNIT=="Week" ~ floor(AGE_AMOUNT*7/365),
                       AGE_UNIT=="Month" ~ floor(AGE_AMOUNT*12/365),
                       AGE_UNIT=="Year" ~ floor(AGE_AMOUNT),
                       TRUE ~ NA_integer_) |>
           coalesce(YEAR_OF_BIRTH)) |>
  mutate(AGE=case_when(AGE>1000L ~ 1911L-AGE,  TRUE ~ AGE)) |>
  count(AGE,SEX) |>
  filter(!is.na(AGE), AGE != 1911L, SEX %in% c("Male","Female")) |>
  mutate(n=if_else(SEX=="Female",-n,n)) |>
  collect() |>
  ggplot(aes(y=factor(AGE),x=n*20,fill=SEX)) +
  geom_col() +
  scale_y_discrete(breaks=seq(0,150,5)) +
  scale_x_continuous(labels=\(x)scales::comma(abs(x))) +
  labs(title="Age pyramid in 1911",
       x="Number of people",y="Age",fill="Sex",
       caption="Canada census 1991 (CCIR 5% samele")
```

The bump for males around the age of 25 is curious, splitting the population into immigrants and non-immigrants reveals that this is caused by selection on males in the immigrant population.

```{r}
census_1911 |>
  mutate(Immigrant = if_else(!is.na(YEAR_OF_IMMIGRATION),"Immigrant","Non-immigrant")) |>
  mutate(AGE=case_when(AGE_UNIT=="Day" ~ floor(AGE_AMOUNT/365),
                       AGE_UNIT=="Week" ~ floor(AGE_AMOUNT*7/365),
                       AGE_UNIT=="Month" ~ floor(AGE_AMOUNT*12/365),
                       AGE_UNIT=="Year" ~ floor(AGE_AMOUNT),
                       TRUE ~ NA_integer_)) |>
  mutate(AGE=case_when(is.na(AGE) & !is.na(YEAR_OF_BIRTH) ~ 1911L-YEAR_OF_BIRTH, 
                       TRUE ~ AGE)) |>
  mutate(AGE=case_when(AGE>1000L ~ 1911L-AGE, 
                       TRUE ~ AGE)) |>
  count(AGE,SEX,Immigrant) |>
  filter(!is.na(AGE), AGE != 1911L, SEX %in% c("Male","Female")) |>
  mutate(n=if_else(SEX=="Female",-n,n)) |>
  collect() |>
  ggplot(aes(y=factor(AGE),x=n*20,fill=SEX)) +
  geom_col() +
  facet_wrap(~Immigrant) +
  scale_y_discrete(breaks=seq(0,150,5)) +
  scale_x_continuous(labels=\(x)scales::comma(abs(x))) +
  labs(title="Age pyramid in 1911",
       x="Number of people",y="Age",fill="Sex",
       caption="Canada census 1991 (CCIR 5% samele")
```

Checking in by place of birth in the immigrant population, just taking the top 9 immigrant countries, shows that immigrants from all countries tended to be male-heavy with total volumne clusterd around the 25 year age group. For Italy this pattern was particularly strong.

```{r}
top_births <- census_1911 |> 
  filter(!is.na(YEAR_OF_IMMIGRATION)) |>
  count(INDIVIDUAL_BIRTH_COUNTRY) |>
  slice_max(n=9,order_by = n) |>
  arrange(-n) |>
  pull(INDIVIDUAL_BIRTH_COUNTRY)

census_1911 |>
  filter(INDIVIDUAL_BIRTH_COUNTRY %in% top_births) |>
  mutate(AGE=case_when(AGE_UNIT=="Day" ~ floor(AGE_AMOUNT/365),
                       AGE_UNIT=="Week" ~ floor(AGE_AMOUNT*7/365),
                       AGE_UNIT=="Month" ~ floor(AGE_AMOUNT*12/365),
                       AGE_UNIT=="Year" ~ floor(AGE_AMOUNT),
                       TRUE ~ NA_integer_)) |>
  mutate(AGE=case_when(is.na(AGE) & !is.na(YEAR_OF_BIRTH) ~ 1911L-YEAR_OF_BIRTH, 
                       TRUE ~ AGE)) |>
  mutate(AGE=case_when(AGE>1000L ~ 1911L-AGE, 
                       TRUE ~ AGE)) |>
  count(AGE,SEX,INDIVIDUAL_BIRTH_COUNTRY) |>
  filter(!is.na(AGE), AGE != 1911L, SEX %in% c("Male","Female")) |>
  mutate(n=if_else(SEX=="Female",-n,n)) |>
  collect() |>
  mutate(INDIVIDUAL_BIRTH_COUNTRY=factor(INDIVIDUAL_BIRTH_COUNTRY,
                                         levels=top_births)) |>
  ggplot(aes(y=factor(AGE),x=n*20,fill=SEX)) +
  geom_col() +
  facet_wrap(~INDIVIDUAL_BIRTH_COUNTRY,scales="free_x") +
  scale_y_discrete(breaks=seq(0,150,10s)) +
  scale_x_continuous(labels=\(x)scales::comma(abs(x))) +
  labs(title="Age pyramid in 1911",
       x="Number of people",y="Age",fill="Sex",
       caption="Canada census 1991 (CCIR 5% samele")
```


```{r}
close_pumf(census_1911)
```

