---
title: "canpumf Pipeline Architecture"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{canpumf Pipeline Architecture}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

This document describes how `get_pumf()` turns a Statistics Canada PUMF zip
file into a lazy DuckDB-backed `dplyr::tbl()`, covering every choice and
fallback along the way. The LFS has its own accumulating pipeline, described
separately at the end.

---

## High-level flow

```{r pipeline-diagram, echo=FALSE, out.width="100%", fig.alt="canpumf pipeline: get_pumf dispatches LFS vs. the three-stage pipeline (locate/download, parse metadata, build DuckDB), then registers provenance and returns a lazy tbl."}
# The diagram is rendered to a static image (SVG for HTML, PNG for PDF) via
# Graphviz so it renders identically and reliably in every output format,
# without relying on JavaScript htmlwidgets (which never render in PDF).
dot <- '
digraph pipeline {
  graph [rankdir = TB, fontname = "Helvetica", nodesep = 0.30, ranksep = 0.40, compound = true];
  node  [fontname = "Helvetica", fontsize = 10, style = "filled", fillcolor = "#eef3f8", color = "#5b7da3", margin = "0.09,0.05"];
  edge  [fontname = "Helvetica", fontsize = 9, color = "#666666", arrowsize = 0.7];

  A    [label = "get_pumf(series, version, lang)", shape = box, style = "filled,rounded", fillcolor = "#d9ead3"];
  LFS  [label = "series == LFS?", shape = diamond, fillcolor = "#fce8b2"];
  LFSP [label = "lfs_get_pumf()", shape = box, style = "filled,rounded", fillcolor = "#d9ead3"];

  A   -> LFS;
  LFS -> LFSP [label = "yes"];
  LFS -> CHK  [label = "no", lhead = cluster_s1];

  subgraph cluster_s1 {
    label = "Stage 1  —  locate / download";
    labeljust = "l"; fontname = "Helvetica-Bold"; fontsize = 11;
    style = "rounded,filled"; fillcolor = "#fbfdff"; color = "#9fb6cc";
    CHK  [label = "version dir exists?", shape = diamond, fillcolor = "#fce8b2"];
    COL  [label = "look up collection URL", shape = box];
    EFT  [label = "EFT-only?", shape = diamond, fillcolor = "#fce8b2"];
    ERR  [label = "stop: deposit zip manually", shape = box, fillcolor = "#f4cccc"];
    DL   [label = "download zip", shape = box];
    UZ   [label = "robust_unzip()", shape = box];
    EXTR [label = "zip already extracted?", shape = diamond, fillcolor = "#fce8b2"];

    CHK  -> EXTR [label = "yes, not refresh"];
    CHK  -> COL  [label = "no / refresh"];
    COL  -> EFT;
    EFT  -> ERR  [label = "yes"];
    EFT  -> DL   [label = "no"];
    DL   -> UZ;
    UZ   -> EXTR;
    EXTR -> UZ   [label = "no"];
  }

  subgraph cluster_s2 {
    label = "Stage 2  —  parse metadata";
    labeljust = "l"; fontname = "Helvetica-Bold"; fontsize = 11;
    style = "rounded,filled"; fillcolor = "#fbfdff"; color = "#9fb6cc";
    MC   [label = "metadata already exists?", shape = diamond, fillcolor = "#fce8b2"];
    DF   [label = "detect_formats()", shape = box];
    P1   [label = "LFS codebook.csv", shape = box];
    P2   [label = "CPSS variables.csv", shape = box];
    P3   [label = "SAS cards (.lay + .lbe)", shape = box];
    P4   [label = "SPSS split (vare/vale/_i)", shape = box];
    P5   [label = "SPSS mono (.sps / SPSS.txt / .xmf)", shape = box];
    P6   [label = "SPSS .sav", shape = box];
    P7   [label = "PDF Dictionary", shape = box];
    MRG  [label = "merge_metadata()", shape = box];
    WR   [label = "write variables.csv / codes.csv / layout.csv", shape = box];

    MC -> DF [label = "no / refresh"];
    DF -> P1; DF -> P2; DF -> P3; DF -> P4; DF -> P5; DF -> P6; DF -> P7;
    P1 -> MRG; P2 -> MRG; P3 -> MRG; P4 -> MRG; P5 -> MRG; P6 -> MRG; P7 -> MRG;
    MRG -> WR;
  }

  EXTR -> MC [label = "yes", lhead = cluster_s2];

  subgraph cluster_s3 {
    label = "Stage 3  —  build DuckDB";
    labeljust = "l"; fontname = "Helvetica-Bold"; fontsize = 11;
    style = "rounded,filled"; fillcolor = "#fbfdff"; color = "#9fb6cc";
    TB   [label = "table already in DuckDB?", shape = diamond, fillcolor = "#fce8b2"];
    FF   [label = "find data file", shape = box];
    FWF  [label = "layout.csv exists\nand file not .csv?", shape = diamond, fillcolor = "#fce8b2"];
    RFW  [label = "read_fwf", shape = box];
    RCS  [label = "read_csv", shape = box];
    JNK  [label = "drop trailing junk rows", shape = box];
    FX   [label = "apply data fixups\n(str_pad, rename, cols_swap, force_*)", shape = box];
    BSW  [label = "BSW mask in registry?", shape = diamond, fillcolor = "#fce8b2"];
    RBW  [label = "join bootstrap weights", shape = box];
    NC   [label = "numeric conversion\n(missing ranges + na_values)", shape = box];
    CL   [label = "code labels to factors", shape = box];
    WD   [label = "write DuckDB table", shape = box];
    EN   [label = "enforce ENUM / force_* types", shape = box];
    OD   [label = "open read-only connection", shape = box];

    TB  -> FF  [label = "no / refresh"];
    FF  -> FWF;
    FWF -> RFW [label = "yes (FWF)"];
    FWF -> RCS [label = "no (CSV)"];
    RFW -> JNK; RCS -> JNK;
    JNK -> FX;
    FX  -> BSW;
    BSW -> RBW [label = "yes"];
    BSW -> NC  [label = "no"];
    RBW -> NC;
    NC  -> CL;
    CL  -> WD;
    WD  -> EN;
    EN  -> OD;
  }

  WR -> TB  [lhead = cluster_s3];
  MC -> TB  [label = "yes, not refresh"];
  TB -> OD  [label = "yes, not refresh"];

  REG [label = "register provenance (series, version, lang)", shape = box, style = "filled,rounded", fillcolor = "#d9ead3"];
  TBL [label = "return lazy dplyr::tbl()", shape = box, style = "filled,rounded", fillcolor = "#d9ead3"];
  OD  -> REG;
  REG -> TBL;
}
'

have_render <- requireNamespace("DiagrammeR", quietly = TRUE) &&
  requireNamespace("DiagrammeRsvg", quietly = TRUE)

if (have_render) {
  svg <- DiagrammeRsvg::export_svg(DiagrammeR::grViz(dot))
  if (knitr::is_latex_output()) {
    if (requireNamespace("rsvg", quietly = TRUE)) {
      png <- knitr::fig_path(".png")
      dir.create(dirname(png), recursive = TRUE, showWarnings = FALSE)
      rsvg::rsvg_png(charToRaw(svg), png, width = 1800)
      knitr::include_graphics(png)
    } else {
      message("Install rsvg to render the pipeline diagram in PDF output.")
    }
  } else {
    svg_file <- knitr::fig_path(".svg")
    dir.create(dirname(svg_file), recursive = TRUE, showWarnings = FALSE)
    writeLines(svg, svg_file)
    knitr::include_graphics(svg_file)
  }
} else {
  cat("Install DiagrammeR and DiagrammeRsvg to render this diagram.")
}
```

---

## Stage 1 — Locate or download

`pumf_locate_or_download()` ensures the version directory exists with extracted
content before any parsing begins.

**Cache layout:**

```
<cache_path>/
  <series>/
    <version>/
      <original>.zip          # retained after extraction
      <extracted dirs>/
      metadata/               # written by Stage 2
      <series>_<version>.duckdb
```

**Decision sequence:**

1. **`refresh = TRUE`** — delete the `.duckdb` file(s) and `metadata/`
   subdirectory, leaving the zip and extracted content untouched.  Stages 2 and
   3 then re-run without re-downloading.

2. **`redownload = TRUE`** — wipe the *entire* version directory first, then
   proceed as a first-time run.  Implies `refresh`.

3. **Already extracted?** — `version_is_extracted()` returns `TRUE` if any
   subdirectory (other than `metadata/`) or non-zip non-duckdb file is present.
   If `TRUE`, the zip step is skipped even when the zip is still on disk.

4. **Download** — the URL is looked up in `list_canpumf_collection()`.  Surveys
   distributed only via Statistics Canada's EFT portal have the marker `"(EFT)"`
   instead of a URL; the function stops with instructions to deposit the zip
   manually.

5. **Extract** — `robust_unzip()` handles two edge cases:
   - *Naming collision*: some zips contain a single top-level directory with the
     same name as the archive (e.g. `2025-CSV.zip/`).  The colliding directory
     is renamed to strip `.zip` before being moved into the version directory.
   - *Encoding*: older StatCan zips store filenames without the UTF-8 flag
     (General Purpose Bit Flag bit 11).  `grep`/`sub` calls on zip entry names
     use `useBytes = TRUE` to avoid "invalid in this locale" warnings.

### Version resolution

`pumf_resolve_version()` canonicalises Census version strings before any
registry lookup.  Any string starting with a four-digit year is parsed
flexibly: the file type is detected by grepping for `"hierarchical"`,
`"household"`, or `"famil"` (defaulting to `"individuals"`), and CMA vs
provincial by grepping for `"cma"`.  The registry is then probed to determine
the correct canonical format for that year.

Examples:

| User input | Resolved |
|---|---|
| `"2021"` | `"2021 (individuals)"` |
| `"1971"` | `"1971/individuals_prov"` |
| `"1971 CMA"` | `"1971/individuals_cma"` |
| `"1971 households CMA"` | `"1971/households_cma"` |
| `"1986 families"` | `"1986/families"` |
| `"2001 households"` | `"2001 (households)"` |

---

## Stage 2 — Parse metadata

`pumf_parse_metadata()` converts raw SPSS/SAS command files into three
canonical CSVs.  The function is idempotent: it does nothing if
`metadata/variables.csv` already exists and `refresh = FALSE`.

### Format detection

`detect_formats()` scans the entire version directory recursively and
identifies which parser(s) apply.  **Multiple parsers can fire for the same
survey** (e.g. SPSS split for layout/codes and SAS cards for BSW weights).

| Priority | Format | Detection rule |
|---|---|---|
| 1 | **LFS codebook CSV** | filename matches `codebook\.csv` (case-insensitive) |
| 2 | **CPSS variables CSV** | filename is exactly `variables.csv` |
| 3 | **SAS reading cards** | directory contains both a `.lay` and a `.lbe` file |
| 4 | **SPSS split-file** | any `.sps` file whose name ends in `vare`, `vale`, or `_i` |
| 5 | **SPSS monolithic** | `.sps` file, `*SPSS.txt` file, or `.xmf` file whose content contains `VALUE LABELS` **or `DATA LIST`** (checked with `useBytes = TRUE` to tolerate CP850/Latin-1 data); `VARIABLE LABELS` is optional |
| 6 | **SPSS `.sav`** | a `.sav` binary file readable by haven |
| 7 | **PDF Data Dictionary** | `*Dictionary.pdf` present and `pdftools` installed; supplements label-only surveys where the SPSS file has `DATA LIST` but no `VARIABLE LABELS` or `VALUE LABELS` |

Detection for case 5 also searches for a parallel French file — any candidate
in the same set whose path includes `/fran` or `/french` (case-insensitive).

### Parsers

#### SPSS monolithic (`parse_spss_mono`)

Handles the single-file SPSS format used by Census (2001–2021), SFS 1999,
SHS, and others.  The file typically contains `DATA LIST`, `VARIABLE LABELS`,
`VALUE LABELS`, and sometimes `MISSING VALUES` and `FORMATS` sections.
`VARIABLE LABELS` is optional (e.g. Census 2011 individuals omits it).
Older releases like SFS 1999 have only `DATA LIST` with no label sections at
all — these produce a fully importable table with raw codes but no
human-readable factor levels.

Key parsing details:

- **Column ranges** — `DATA LIST` ranges may have spaces on either side of the
  dash (`129-135`, `129 - 135`, or `129-  135`).  All three are normalised by
  the regex `(\\d+)\\s*-\\s*(\\d+)` before tokenisation.

- **Record-group marker** — a leading `/` on the first variable line (e.g.
  `/PROVP  1-2`) is stripped, not discarded, so the variable is retained.

- **Section terminator** — the `DATA LIST` section ends at the first blank
  line, `.` line, or occurrence of `VARIABLE LABELS`, `VALUE LABELS`,
  `MISSING VALUES`, `FORMATS`, or `EXECUTE` at the start of a line.  The
  keyword check is the reliable terminator for older files (e.g. 1991 XMF)
  that have no blank line between `DATA LIST` and `VARIABLE LABELS`.

- **DATA LIST type annotations** — the `(A)` suffix after a column range marks
  a character-type variable.  The parser records an `is_char` flag per column
  and uses it to populate `variables.csv` types when no `VARIABLE LABELS`
  section is present.

- **Sentinel detection** — variables whose only VALUE LABELS are sentinel
  phrases ("Not applicable", "Valid skip", "Don't know", "Data not available",
  etc.) are classified as `numeric` with a `missing_low/missing_high` range,
  not as `character`.  This prevents spurious NA warnings when numeric values
  fall outside the label set.

- **Zero-padded codes** — unquoted SPSS numeric codes like `01`, `02` are
  normalised via `as.numeric()` → `as.character()` so they match bare integer
  values in CSV data.

- **Multi-variable VALUE LABELS blocks** — `/VAR1 VAR2 VAR3` headers
  (possibly spanning continuation lines) are fully parsed so all listed
  variables receive the code/label pairs.

#### SPSS split-file (`parse_spss_split`)

Used by SFS, CPSS, and similar surveys that ship separate files for variable
labels (`*vare.sps`), value labels (`*vale.sps`), missing values
(`*miss.sps`), and layout (`*_i.sps`).  The `layout_mask` from the registry
disambiguates when a single directory holds multiple sets (e.g. individual vs.
household files).

#### SAS reading cards (`parse_sas_cards`)

`.lay` files supply the fixed-width column positions; `.lbe` files supply the
value labels in `PROC FORMAT` syntax.  Variable labels come from a companion
`.sas` file if present.  This parser reuses `parse_spss_split`'s layout parser
since the `.lay` format is identical.

#### LFS codebook CSV (`parse_lfs_codebook`)

The LFS ships a single `*codebook.csv` with one row per code value.  Columns
are always read as CP1252 regardless of the `metadata_encoding` registry field.

#### CPSS variables CSV (`parse_cpss_csv`)

The Canadian Perspectives Survey Series ships a `variables.csv` with variable
metadata only; no layout or codes.  The encoding defaults to Latin-1 (CP1252
if the registry overrides).

#### SPSS `.sav` (`parse_spss_sav`)

Haven is used for binary `.sav` files when no text-format command file is
available.  This is a fallback for surveys that do not ship SPSS syntax.

#### PDF Data Dictionary (`parse_pdf_dictionary`)

StatCan PDF Data Dictionaries follow a standard bilingual format.  Variable
blocks start with `<name>  Position: N  Character/Numeric(w)`.  The parser
extracts variable long-names (`Long name:` / `Long nom:`) and code-value
labels (`Codes:` / `Domaine:`).  Reserved codes (`Reserved Codes:` /
`Codes Réservés:`) set `missing_low/missing_high` ranges.

This parser produces only `variables` and `codes` (no `layout`), and fires
only when `pdftools` is installed and a matching `*Dictionary.pdf` is found.
It is used as a label-only supplement for surveys like SFS 1999 where the SPSS
file is `DATA LIST`-only.

### Metadata encoding

The registry `metadata_encoding` field sets the encoding for all text-format
parsers.  Default is `"CP1252"` (a superset of Latin-1 that correctly decodes
Windows-era en-dashes and curly quotes).  Exceptions:

| Surveys | Encoding | Reason |
|---|---|---|
| Census 2021, 2021 hierarchical | `"UTF-8"` | Command files shipped as UTF-8 |
| Census 1991 (individuals) | `"CP850"` | DOS-era IBM Code Page 850 |

### Merge

`merge_metadata()` takes the list of parser outputs and produces a single
`list(variables, codes, layout)`.  Conflicts are resolved: later parsers win
on duplicate variable names.  If a layout is present in only some parsers, the
function checks that every variable with a layout entry also appears in
`variables`, stopping with a diagnostic otherwise.

The final result is written to:

- `metadata/variables.csv` — one row per variable (name, label_en, label_fr,
  type, decimals, missing_low, missing_high)
- `metadata/codes.csv` — one row per code value (name, val, label_en,
  label_fr)
- `metadata/layout.csv` — one row per fixed-width column (name, start, end);
  absent for CSV-format surveys

---

## Stage 3 — Build DuckDB

`pumf_build_duckdb()` reads the canonical CSVs from `metadata/`, reads the
raw data file, applies transformations, and writes a `.duckdb` file.  The
function skips the build if the target table already exists and `refresh =
FALSE`.

### Data file selection

`find_pumf_data_file()` searches the version directory recursively.

**Extension pre-filter** — derived from the registry `file_mask`:

| `file_mask` ends in | Pre-filter |
|---|---|
| `.csv` | only files matching `\.csv$` |
| `.txt` or `.dat` | only files matching `\.(txt\|dat)$` |
| other / unusual (e.g. `.INDIV`) | all files (relies on `file_mask` alone) |
| absent + layout exists | `\.(txt\|dat)$` (FWF inferred from layout) |
| absent + no layout | `\.csv$` |

Several subdirectories are always excluded from the search: `metadata/`,
`SPSS/`, `Command/`, `Syntax/`, `Layout/`, `SpssCard/`, `Reading_cards/`,
`Documents/`.  Bootstrap weight (`_BSW.`) files are also excluded; they are
handled separately.

When multiple candidates survive, the `file_mask` regex narrows the list.  If
more than one still remains, the function stops with a message listing the
ambiguous files and asks to set `file_mask` in the registry.

### FWF vs. CSV

After the data file is identified:

- **FWF** when `metadata/layout.csv` exists *and* the data file does not end
  in `.csv`.  This handles the edge case (e.g. CHS) where the SPSS DATA LIST
  produces a layout but the actual data ships as CSV.
- **CSV** otherwise.

Both paths read all columns as character (`col_types = cols(.default = "c")`)
to preserve leading zeros and avoid premature type coercion.  Numeric
conversion happens explicitly in the next step.

### Trailing junk row removal (FWF only)

After reading a fixed-width file, any row where fewer than two columns are
non-NA is dropped.  FWF files from older StatCan archives often end with
`\r\n\x1a` (a DOS EOF marker), which the FWF reader interprets as a
one-character row with a single non-NA field; this step removes it silently.
CSV files are not affected — CSV parsers handle trailing newlines correctly.

### Data fixups (pre-label)

Registry `data_fixups` entries are applied to the raw character data before
label mapping:

- **`str_pad`** — left- or right-pad specified columns to a target width.  Used
  to zero-pad codes that arrive without leading zeros in some CSV formats (e.g.
  SFS).
- **`rename`** — rename a column; applied only when the old name is present
  (safe for surveys that ship in multiple release variants, e.g. Census 2021
  RELIG/RELIGION_DER).
- **`cols_swap`** — named character vector `c(A = "B", C = "D")` swapping
  pairs of column names.  Used for surveys where the DATA LIST variable names
  are transposed relative to the PDF documentation (e.g. WKACTMA/WKACTFA and
  FAOCC81/MAOCC81 in Census 1981 individuals).
- **`force_numeric`** — character vector of column names to treat as numeric
  regardless of how many VALUE LABELS are declared.  Used when a variable
  carries boundary or top-code labels (e.g. `"85 years and over"`) alongside
  otherwise-continuous values, or is an integer index the SPSS file
  mis-classifies as categorical (e.g. SUBSAMPL in Census 1971).  The codes are
  dropped, but any **true-missing** sentinel codes (Not stated, Don't know,
  Valid skip, … — *not* zero-value labels like "None") are first converted into
  a per-variable `missing_low/missing_high` range so those sentinels still
  become `NA`.  An existing missing range (from `MISSING VALUES` or a split-SPSS
  miss file) takes precedence.
- **`force_character` / `force_integer` / `force_bigint`** — character vectors
  of variable names whose **DuckDB storage type** is overridden.  Unlike the
  conversions above, the raw string values are kept verbatim (no numeric
  conversion, no code labeling), so geographic codes retain leading zeros and
  out-of-`int`-range IDs survive.  `force_character` keeps the column VARCHAR;
  `force_integer` / `force_bigint` cast it to INTEGER / BIGINT via
  `ALTER COLUMN` *after* the table is written (an INTEGER cast that overflows
  2^31 errors — use `force_bigint`).  A variable may appear in at most one
  `force_*` set (including `force_numeric`); this is validated at build time.
  LFS sources its `SURVYEAR` / `SURVMNTH` / `REC_NUM` integer-forcing through
  this mechanism from the shared LFS registry entry.
- **`codes_supplement`** — named list of `data.frame`s injecting code-label
  rows absent from the SPSS command files (values present in the data but not
  declared in the command files, e.g. the CHS `PPROV` territories code).  Each
  data frame has columns `val`, `label_en`, `label_fr`.  Setting `label_en = NA`
  marks a value as intentionally missing (produces a silent `NA` factor entry
  without a warning, and without introducing a spurious factor level).  All
  entries are verified in the override ledger
  (`tests/testthat/override_verification.csv`).
- **`na_values`** — character vector of raw string sentinels that become `NA`.
  In numeric columns they are exact-matched and NA'd during numeric conversion;
  in labeled (factor) columns they are silently blanked.  Used for undeclared
  Census income sentinels and SAS-style `"."` missing markers.

### Bootstrap weight join (BSW)

When the registry has `bsw_mask` + `bsw_join_key` + `bsw_file_mask`, the BSW
file is found, read (CSV or FWF), and left-joined onto the main data by the
join key before numeric conversion.

### Numeric conversion

`apply_numeric_conversion()` converts character columns typed `"numeric"` in
`variables.csv`:

1. `as.numeric()` on the raw character values.
2. **Missing range** — values in `[missing_low, missing_high]` become `NA`.
   This handles SPSS-declared `MISSING VALUES` blocks.
3. **`na_values` fixup** — additional raw string sentinels from the registry
   (`data_fixups$na_values`) are set to `NA` via `trimws(raw) %in% na_values`.
   Used for undeclared income sentinels in older Census files.

The two mechanisms complement each other: the SPSS `MISSING VALUES` block
catches sentinels declared in the command file; `na_values` catches those that
StatCan omitted from the command file but documents in the user guide.

**Census income sentinel widths** (confirmed from SPSS DATA LIST sections):

| Census years | Income field width | Sentinels (`na_values`) |
|---|---|---|
| 2016, 2021 | 8 chars | `"99999999"`, `"88888888"` |
| 1991–2011 | 7 chars | `"9999999"`, `"8888888"` |
| 1986 and earlier | unverified | none applied |

The two sets are kept separate: applying the 7-digit sentinel to an 8-char
field would incorrectly NA out a valid $9,999,999 income value (stored as
`" 9999999"` which trims to `"9999999"`).

### Code labels → factors

`apply_code_labels()` maps raw character values to R factors using
`codes.csv`.  The factor levels are the complete ordered set from the codes
table, not just the values present in the data.

Unmatched raw values become `NA` with a warning showing the first five
offending values.  An exception is made for values that appear in `codes.csv`
with `label_en = NA` (injected via `codes_supplement`): these are treated as
*intentionally* NA and silently produce `NA` factor entries without a warning.

When `lang = "fra"`, any missing French label falls back per-row to the
English label.

### DuckDB write and ENUM enforcement

The labelled data frame is written to DuckDB with `dbWriteTable()`.  Factor
columns are stored as DuckDB `ENUM` types.  DuckDB >= 1.5.2 does this
automatically; for older versions `ensure_enum_columns()` runs
`ALTER TABLE ... ALTER COLUMN ... TYPE ENUM(...)` for each factor column.

A separate DuckDB table is created per language (table names `eng` and `fra`,
or `eng_<layout_mask>` for surveys with multiple file types).  The write
connection is shut down before `pumf_open_duckdb()` re-opens the file in
read-only mode, preventing in-process lock conflicts when building both
language tables in the same session.

---

## LFS pipeline

The LFS is handled by `lfs_get_pumf()` (delegated directly from `get_pumf()`
without going through `get_pumf_connection()`).  Instead of one DuckDB per
version, all LFS versions share a single `<cache_path>/LFS/LFS.duckdb` with
accumulating tables `lfs_eng` and `lfs_fra`.

Key differences from the standard pipeline:

- **Schema evolution** — when a new LFS version adds a variable absent from
  earlier versions, the column is added via `ALTER TABLE ADD COLUMN`.  When a
  variable changes type (e.g. VARCHAR → ENUM), `ALTER COLUMN SET DATA TYPE`
  is used.

- **Annual supersedes monthly** — if annual and monthly versions for the same
  year are both loaded, the annual version supersedes the monthly rows for that
  year.

- **Version tracking** — a `lfs_versions` table in the shared DuckDB records
  which versions have been downloaded and parsed, so `refresh = "auto"`
  downloads only new versions.

- **Read-only fast path** — when the requested version is already in the
  database, `lfs_get_pumf()` opens only a read-only connection and returns
  immediately.  No write lock is acquired unless new data actually needs to be
  written.

- **`get_pumf()` return** — when a specific version is requested, the function
  applies a `dplyr::filter()` on `SURVYEAR` (and `SURVMNTH` for monthly
  requests) over the full shared table.  Calling `get_pumf("LFS")` without a
  version returns the unfiltered table.

- **`label_pumf_columns()` for LFS** — because the shared schema is the union
  of all loaded versions, variables introduced in later years (e.g. `GENDER`
  added ~2020) are absent from older versions' `variables.csv`.
  `label_pumf_columns()` therefore reads and merges metadata from *every*
  loaded version directory in chronological order, with the most-recent label
  winning on conflicts.

---

## Connection provenance registry

`get_pumf()` registers `(series, version, cache_path, lang)` in a
package-level environment keyed by the DuckDB connection's C++ external-pointer
address:

```
.pumf_con_registry  <- new.env(hash = TRUE, parent = emptyenv())
key = format(con@conn_ref)   # stable across R-level S4 copies
```

This key survives `dplyr` tbl transformations and `select()`/`filter()` calls
because those operations do not create new connections.  `label_pumf_columns()`
uses `.pumf_lookup_con()` to retrieve the provenance; `close_pumf()` removes
the entry and disconnects.

This internal provenance registry is distinct from the **RStudio Connections
pane**.  Whether the DuckDB connection is advertised to that pane is controlled
separately by the `register_connection` argument to `get_pumf()` (default
`getOption("canpumf.register_connection", TRUE)`); set it to `FALSE` to keep
the pane from being spammed when opening and closing many connections
programmatically.

---

## Registry configuration

`pumf_registry_lookup(series, version)` returns a named list that controls
every per-survey choice in the pipeline.  Surveys without an entry use
auto-detection with defaults (see *Newest-sibling inheritance* below for the
one exception).

| Field | Purpose | Default |
|---|---|---|
| `file_mask` | regex to select the data file | `NULL` (auto) |
| `layout_mask` | SPSS file disambiguator for split-file surveys | `NULL` |
| `data_encoding` | encoding of the raw data file | `"CP1252"` |
| `metadata_encoding` | encoding of SPSS/SAS command files | `"CP1252"` |
| `bsw_mask` | `layout_mask` for BSW-specific SPSS files | `NULL` |
| `bsw_file_mask` | filename pattern for the BSW data file | `NULL` |
| `bsw_join_key` | column(s) to join BSW onto the main data | `NULL` |
| `bsw_drop_cols` | BSW columns to drop before joining | `character(0)` |
| `data_fixups` | list of `str_pad`, `rename`, `cols_swap`, `force_numeric`, `force_character`, `force_integer`, `force_bigint`, `codes_supplement`, `na_values` transforms | `list()` |
| `missing_supplement` | named list of `c(lo, hi)` pairs — explicit missing-range overrides for sentinels no generic pattern can classify (e.g. non-integer sentinels like `999.5`) | `NULL` |
| `doc_mask` | regex applied to PDF filenames to filter a shared documentation directory to the relevant file type (e.g. `"Family\|Familles"` for 1986 Census families) | `NULL` |

### Newest-sibling inheritance

Surveys without a registry entry normally fall back to pure auto-detection,
with one exception.  When the requested version is a bare four-digit year and
the same series already has at least one other year-keyed entry,
`pumf_registry_lookup()` inherits the configuration of the newest registered
sibling whose year is <= the requested year (or the oldest sibling if the
requested year predates them all).  This lets a freshly released year deposited
in the cache reuse the prior year's config — which works cleanly now that recent
`file_mask`s use a generic `\d{4}` year placeholder rather than a hard-coded
year.

A `message()` fires once per session so the implicit reuse is discoverable; a
genuinely changed release (new file layout, codes, or BSW join) still needs its
own explicit entry.  Inheritance is **skipped** for multi-part versions (e.g.
Census `2021 (individuals)`) and for LFS, which has its own shared registry
entry.
| `doc_mask` | regex applied to PDF filenames to filter a shared documentation directory to the relevant file type (e.g. `"Family\|Familles"` for 1986 Census families) | `NULL` |
